Risk factors for severe COVID-19 in Scotland: a population-based case-control study

### 2020-05-06 casecontrol.Rmd
a few edits

---
title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'

author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.

@@

pvalue.latex <- function(x, n=1, nexp=1) {
  sapply(x, function(z) {
    if (is.na(z)) {
        return(NA)
    } else if(z > 0.001) {
        return(signif(z, 1))
    } else {
    z <- sprintf("%.*E", 0, signif(z, n)) # default is 1 sig fig
#    z <- sprintf("%.*E", n, signif(z, n + 1))
    z <- as.numeric(unlist(strsplit(z, "E"))) # split z at E
    sprintf("\\ensuremath{%.*f\\times 10^{%0*d}}", 0, z[1], nexp, z[2])
                                        #    sprintf("\\ensuremath{%.*f\\times 10^{%0*d}}", n, z[1], nexp + 1, z[2])
    }
  })
}

## create and format a 95% confidence interval around an odds/hazard ratio
or.ci <- function(coeff, se, ndigits=2) {
  ci.lo <- exp(coeff - 1.96 * se)
  ci.up <- exp(coeff + 1.96 * se)
  sprintf("%.*f (%.*f, %.*f)", ndigits, exp(coeff), ndigits, ci.lo, ndigits, ci.up)
}

```
\newpage

# Background
Case series data from hospitals has suggested a preponderance of certain conditions including diabetes and cardiovascular disease, and elevated risks in those of non-European ethnic origin. Quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator. Furthermore it has not been possible to disentangle whether drug effects may underlie the apparent association with certain co-morbid groups.

In Scotland the existence of a laboratory database with nationwide coverage and a unique health care identifier on all health records has enabled us to ascertain all those with tests positive for SARS-CoV-2 and link to other health records.  Here we describe a population wide case-control study of test confirmed COVID-19 disease in Scotland. From this we report the demographic factors, co-morbidities and concurrent medications associated with being hospitalised for COVID-19 and for becoming a severe COVID-19 case i.e. either warranting critical care support or dying within 28 days of a positive test. This study is intended to inform policy on defining vulnerable groups, identify any drug classes that influence risk of severe disease, and lay a basis for evaluating the feasibility of risk-stratified control measures.

# Methods
## Data sources 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally.  .This allowed all those with a positive viral nucleic acid test for SARS-CoV-2 to be ascertained, hereafter denoted as test positive cases.  At the time writing, serology testing results for auto-antibodies indicating prior infection with SARS-CoV-2 are not yet available.

The National Community Health Index (CHI) database assigns a unique health care identifier to everyone in the population, and is used in all care systems in Scotland.  Using the CHI identifier, records from ECOSS were linked to hospitalization and discharge data from the weekly updated Scottish Morbidity Record 01 and the daily updated RAPID-24 databases. Admission to critical care was ascertained from the national audit database of critical care units (SICSAG). Death registrations and cause of death were obtained from linkage to the National Register of Scotland.  All these databases have nationwide coverage. At the time of writing the national policy on testing has evolved since the start of the epidemic with testing initially being restricted to possible cases that required hospital admission but widening to a "four-pillar" approach that has increased testing in health care workers residential care home settings. Accordingly the probability of being tested has changed over time.  For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case. Note that this is an incidence density sampling design i.e. a control who becomes a case at a later point is retained in the control set. With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.

For all cases and controls co-morbid conditions, the Charlson Comorbidity Index and drug exposure data were available from i) a five year look back query of the Scottish Morbidity Record 01 for sets of International Classification of Disease (ICD)-10 codes for various conditions ii) querying the national prescribing records for exposure to various drug classes iii) linkage to the national diabetes register SCI-diabetes.

We initially examined history of a specific list of conditions and drug classes that have been advised as high risk conditions for COVID-19 by Public Health England, World Health organisation or the US Centres for Disease Control. ICD-10 codes for the conditions are listed in Supplementary Materials.

Ethnicity was ascertained from ethnic group coding on prior SMR records and from applying the ONOMAP algorithm to names available in the CHI database. All cases and controls were assigned to a quintile of the Scottish Index of Multiple Deprivation- this is a residential postcode based indicator of socioeconomic status.

## Case definition
We classified test-positive cases into three groups:

1. Severe or fatal COVID-19, defined by entering critical care in the Scottish Intensive Care Society and Audit Group database (SICSAG) OR death within 28 days of a positive nucleic acid test (regardless of the cause of death given on the death certificate).

@@

3. Not hospitalized

The focus on severe disease is relevant to policy makers because decisions about control measures have been based on projections of the requirement for critical care. By restricting the case definition to this group, which combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of test-positive cases, whatever triage procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.

The proportion of fatal cases that were not tested can be examined in national mortality data ...

# Statistical methods
We used  the R statistical package. We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)

We tabulated the prevalence of co-morbidities and current drug use by case status.  We used conditional logistic regression to estimate the univariate odds associated with each covariate of interest conditioned on age sex and GP practice. 
We also report a conditional logistic regression with all covariates entered simultaneously.

We tested each effect for interaction with age and sex.

Models were run overall and by age band.

To select variables to be retained in the model, we used backward selection with a termination criterion that the reduction in deviance is less than the value corresponding to a threshold p-value of 0.01 
Data were split into a 70:30 training:test sets. 
We evaluated the performance of the resulting backward selection model for predicting severe COVID-19 disease in the test set. For this we used the expected information for discrimination as described and used previously

### The initial demographic vars of interest are ( give that age and sex is matched)

 ethnic group
 SIMD

### The initial comorbidities of interest are listed below

@@

# Information governance
This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data. Datasets were de-identified before analysis.

@@

## Association of demographic factors with severe disease
Table \ref{tab:demog} shows univariate associations of demographic factors with disease.  Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home (rate ratio 8.6).  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.

@@

pander(table.demog.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       split.cells=c(15, 15, 25, 8),
       caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")

@@

table.stepwise <- as.data.frame(stepwise.all[, c(1, 3, 5)])
colnames(table.stepwise) <- c("Coefficient (log rate ratio)", "Standard error", "p.value")
rownames(table.stepwise) <- replace.names(rownames(table.stepwise))

@@

pander(table.stepwise, table.style="multiline",
       # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
	   digits=c(3, 3, 1), 
	   split.table="Inf",
       caption="\\label{tab:stepwise}Stepwise regression: variables retained in model for severe disease")

@@

## Predictive performance of a risk classifier

The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.  Table \ref{tab:wevid} shows that the C-statistic was `r summary(stepwise.densities)[3]` and the expected information for discrimination was  `r summary(stepwise.densities)[5]` bits.

@@

pander(summary(stepwise.densities), table.style="multiline",
       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )

@@

plotWdists(stepwise.densities) + theme_grey(base_size = 18) + 
	theme(legend.title=element_blank())

@@

>>>>

---
title: 'Risk factors for severe COVID-19 in Scotland: a population-based case-control study'
author: 
  - name: Paul M McKeigue
    email: paul.mckeigue@ed.ac.uk
    affiliation: Usher
  - name: Amanda Weir 
    affiliation: HPS
  - name: Jen Bishop
    affiliation: HPS
  - name: Stuart McGurnaghan
    affiliation: IGMM
  - name: Sharon Kennedy
    affiliation: HPS
  - name: Sharon Hutchinson 
    affiliation: GCU
  - name: David McAllister 
    affiliation: Glasgow
  - name: Chris Robertson 
    affiliation: Strathclyde
  - name: Rachael Wood
    affiliation: ISD
  - name: Nazir Lone
    affiliation: Usher
  - name: Tom Caparrotta 
    affiliation: IGMM
  - name: Helen M Colhoun
    affiliation: IGMM
address: 
   - code: Usher
     address: Usher Institute, College of Medicine and Veterinary Medicine, University of Edinburgh, Teviot Place, Edinburgh EH8 9AG, Scotland. PM - Professor of Genetic Epidemiology and Statistical Genetics. NL - Clinical Senior Lecturer in Critical Care
   - code: IGMM
     address: Institute of Genetics and Molecular Medicine, College of Medicine and Veterinary Medicine, University of Edinburgh, Western General Hospital Campus, Crewe Road, Edinburgh EH4 2XUC, Scotland. HC - Axa Chair in Medical Informatics and Epidemiology. TC - Sir George Alberti Doctoral Fellow in Pharmacoepidemiology. 
   - code: Glasgow
     address: Institute of Health and Wellbeing, University of Glasgow, 1 Lilybank Gardens, Glasgow G12 8RZ. DM - Senior Clinical Lecturer in Public Health 
   - code: HPS
     address: Health Protection Scotland (Public Health Scotland), Meridian Court, 5 Cadogan Street, Glasgow G2 6QE
   - code: Strathclyde
     address: Department of Mathematics and Statistics, University of Strathclyde, 16 Richmond Street, Glasgow G1 1XQ. CR - Professor of Public Health Epidemiology
   - code: GCU
     address: School of Health and Life Sciences, Glasgow Caledonian University. SH - Professor of Epidemiology and Population Health 
   - code: ISD
     address: NHS Information Services Division (Public Health Scotland), Gyle Square, 1 South Gyle Crescent, Edinburgh, EH12 9EB.  RW - Consultant in Maternal and Child Health.

@@

source("helperfunctions.R")

\newpage

# Background
Reports based on cases seen in hospitals hve suggested that in those with severe disease the prevalence of diabetes and cardiovascular disease is higher than expected, and that a higher than expected proportion of severe cases are of non-European ethnic origin. Quantification of the risk associated with these characteristics and co-morbidities has been limited by the lack of an appropriate background population comparator. Furthermore it has not been possible to disentangle whether drug effects may underlie the apparent association with certain co-morbid groups.

The aim of this study was to investigate risk factors for severe or fatal cases of COVID-19 confirmed by testing.  Specific objectives were to test hypotheses about drugs classes associated with altered risk, and to construct a risk prediction algorithm that can be used to identify high-risk individuals who could be offered shielding during the epidemic.  To achieve this aim we conducted a case-control study of all severe and fatal cases in Scotland, with controls matched for age, sex and primary care practice.

# Methods
## Case definition 
The Electronic Communication of Surveillance in Scotland (ECOSS) database captures all virology testing in all NHS laboratories nationally, using the Community Health Index (CHI) identifier.  The CHI number is a unique identifier for everyone in the population, and is used in all care systems in Scotland.  All individuals testing positive for nucleic acid for SARS-CoV-2 up to `r format(max(cc.all$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained.  Admissions to critical care were obtained from the national audit database of the Scottish Intensive Care Society and Audit Group (SICSAG)ical care units (SICSAG). Hospital admissions from the time of testing were obtained from the RAPID-24 database which is updated each day. 
Death registrations up to `r format(max(cc.all$Date.Death, na.rm=TRUE), "%d %B %Y")`
and cause of death were obtained from linkage to the National Register of Scotland.

We classified test-positive cases into three groups:

1. Severe or fatal COVID-19, defined by a record of entering critical care in the SICSAG database, or death within 28 days of a positive nucleic acid test regardless of the cause of death given on the death certificate.

@@

3. Not hospitalised

By restricting the case definition to a group that combines those receiving critical care with those who died without receiving critical care, we ensure complete ascertainment of severe test-positive cases, whatever selection procedures may have been in place to exclude those assessed as unlikely to benefit from critical care.  This focus on severe or fatal disease is relevant to policy because decisions about control measures have been based on projections of the requirement for critical care.

In national mortality data, XX % of those certified with COVID-19 as underlying case of death had a positive test record.

## Matched controls
For each test-positive case, we ascertained ten matched controls of the same sex, one year age band and attending the same General Practice as each case alive on the same day as the date of test in the case. Note that this is an incidence density sampling design i.e. a control who becomes a case at a later point is retained in the control set. With this design, the odds ratios in a conditional logistic regression model are equivalent to rate ratios.

## Demographic data
Ethnicity of cases and controls was classified using a name classification algorithm [@lakha_name_2011] with the names in the CHI database.  This is able to distinguish South Asian and Chinese groups from the White group, but misclassifies Black-Caribbean people (a relatively small group in Scotland) as White.  For those with a hospital admission record, a separate coding of ethnicity was available using self-described categories based on those used in the national census.  Socioeconomic status was scored as the Scottish Index of Multiple Deprivation, based on the postcode in the CHI database.

## Morbidity and drug prescribing

All cases and Using the CHI identifier, records from ECOSS were linked to hospital discharge records in the Morbidity Record 01 (SMR01).  and

For all cases and controls, ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record 01 (SMR01, excluding records of discharges less than 25 days before testing positive for SARS-Cov-2. BNF drug codes were extracted from the last year of encashed prescriptions, excluding those encashed less than 15 days before testing positive for SARS-CoV-2.  Diagnosed cases of diabetes were identified through linkage to the national diabetes register (SCI-Diabetes), with a clinical classification of diabetes type as Type 1, Type 2 or Other.

We initially examined history of a specific list of conditions and drug classes that have been advised as high risk conditions for COVID-19 by Public Health England, World Health organisation or the US Centres for Disease Control.

Co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an encashed prescription of a drug for which that diagnostic code is the only indication.

The ICD-10 codes and drug codes used to define each variable are listed in Supplementary Materials.

## Statistical methods

We calculated the sex and age band specific cumulative incidence of severe COVID-19 disease using the age and sex distribution of those alive in the population as of March 1st 2020. (CHECK)

The univariate association of each variable with severe COVID-19 was modelled with conditional logistic regression, implemented as Cox regression in the R function `survival::clogit`

We tested each effect for interaction with age and sex.  Effect of interest

The analyses were repeated after excluding those aged over 75 years.

To construct a risk prediction model, we used stepwise regression alternating between forward and backward steps, implemented in the R function `stats::step`, so as to maximize the AIC.

The performance of the risk prediction model as a classifier was examined by n-fold cross-validation, with performance calculated over all test folds as the expected information for discrimination and (more conventionally) as the C-statistic.

### Comorbidities of initial interest

@@

## Information governance
This research was conducted under approvals A(i) PBPP approvals for accessing ?????; (ii) existing PBPP approval that allows HPS staff to link any PHI controlled datasets; (janet to add detail and reference numbers?) (iii) PBPP amendment (1617-0147) that covers inward transfer of diabetes data. Datasets were de-identified before analysis.

@@

## Association of demographic factors with severe disease

For the case-control comparison, this version of the manuscript restricts the dataset to those aged less than `r max(cc.severe$AGE) + 1` years.

Table \ref{tab:demog} shows univariate associations of demographic factors with disease.  Higher risk of severe disease was associated with socioeconomic deprivation and with residence in a care home.  In comparison with those classified as White, South Asians and other non-White ethnic minorities were at lower risk of severe disease.

@@

pander(table.demog.aug, 
       digits=c(0, 0, 3, 3, 3, 1),
       round=c(0, 0, 3, 3, 3, 10),
	   split.tables="Inf",
       align=rep("centre", 4), 
       split.cells=c(16, 16, 25, 8),
       caption="\\label{tab:demog}Univariate associations of severe disease with demographic factors")

@@

table.stepwise <- as.data.frame(stepwise.all[, c(1, 5)])
table.stepwise[, 1] <- exp(table.stepwise[, 1])
colnames(table.stepwise) <- c("Rate ratio", "p.value")
rownames(table.stepwise) <- replace.names(rownames(table.stepwise))

@@

pander(table.stepwise, table.style="multiline",
       # split.cells=c(5, 5, 5, 5, 5, 5, 5), 
	   digits=c(3, 1), 
	   split.table="Inf",
       caption="\\label{tab:stepwise}Stepwise regression: variables retained in model for severe disease")

@@

## Predictive performance of a risk classifier
The predictive performance of the model chosen by stepwise regression was estimated by `r nfold`-fold cross-validation, and comparing observed and predicted case status over all test folds.  Table \ref{tab:wevid} shows that the C-statistic was `r summary(all.densities)[3]` and the expected information for discrimination was  `r summary(all.densities)[5]` bits.

@@

pander(summary(all.densities), table.style="multiline",
       split.cells=c(5, 5, 5, 5, 5, 5, 5), split.table="Inf",
       caption="\\label{tab:wevid}Prediction of severe COVID-19: cross-validation of model chosen by stepwise regression" )

@@

plotWdists(all.densities) + theme_grey(base_size = 18) + 
	theme(legend.title=element_blank())

@@

# Discussion

What is the explanation of the 14-fold higher incidence of severe disease in those who have encashed at least one prescription in the last year?

# References {-}

<div id="refs"></div>
